A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-02 tex/codatadp.tex
Refutation soundness proof (draft).

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S^0$ be the set of terms that are arguments of a selector application in $\tEc$.
For $i \geq 0$, let $S^{i+1}$ $=$ $S_i \cup \{ s( t ) \mid t \in S^i, s \in sel( t ), t : \tau \text{ finite} \}$.
Let $S^\ast$ be the fix point of this sequence, which is guarenteed to be finite 
since $S^i \setminus S^0$ only includes selectors applied to terms having finite type.
Now, let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_j}_i( t ) ) \mid t \in S^\ast, C_i \in cons( t ) \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node is 
obtained as a result of adding an equality between two terms from $\tcEc$ to its parent node.
Thus, the maximum depth of a branch in a derivation tree with root node $\Ec$ is $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\end{proof}

\begin{lemma}[Refutation Soundness]
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
%We show that if a derivation tree with $\Ec$ has a model $\M$, 
\end{proof}

>>>>

\begin{lemma}[Termination]
All derivation trees are finite.
\end{lemma}
\begin{proof}
Consider a derivation tree with root node $\Ec$.
Let $S^0$ be the set of terms that are arguments of a selector application in $\tEc$.
For $i \geq 0$, let $S^{i+1}$ $=$ $S_i \cup \{ s( t ) \mid t \in S^i, t : \tau \text{ finite}, s \in \Sel^\tau  \}$.
Let $S^\ast$ be the fixed point of this sequence, which is guarenteed to be a finite set 
since $S^i \setminus S^0$ only includes selectors applied to terms having finite type.
Now, let $\tcEc$ be the set of subterms of $\Ec \cup \{ C_i( s^1_i( t ), \ldots, s^{a_j}_i( t ) ) \mid t \in S^\ast, t : \tau, C_i \in \Ctr^\tau \}$.
In a derivation tree with root node $\Ec$, 
it can be shown by case analysis on the rules of the calculus that each non-root node is 
obtained as a result of adding an equality between two terms from $\tcEc$ to its parent node.
Thus, the maximum depth of a branch in a derivation tree with root node $\Ec$ is $\mid \tcEc \mid^2$,
which is finite since $\tcEc$ is finite.
\qed
\end{proof}

\begin{lemma}[Refutation Soundness]
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
\rem{Elaborate on cyclic}
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since all models of $\thD$ entail exactly one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, or \rn{Bisimilar}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
\rem{Elaborate on Bisimilar}
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and since $\Ec \models_{\thD} t_1 \teq t_2$,
we have that $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}
